---
title: "Mini-Project 01"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Motivations

I work with fuel data weekly, and I get a report from the EIA that had the diesel prices of the week. Working in transportation, fuel is a big part of the business, and clients always want to see how the market is responding to the recent geopolitical events. There were big spikes in response to it, and I would like to see if the fuel market's spikes as per the 1984-2017 fuel economy data produced during vehicle testing at the Environmental Protection Agency’s (EPA) National Vehicle and Fuel Emissions Laboratory also has spikes related to important events. I would imagine so, but: even though that was my original motivation, instead I found that this data does not answer that kind of curiosity, which is *okay*. 

# The Data

```{r}
library(tidyverse) # loading library
fuel <- read_csv("https://raw.githubusercontent.com/aalhamadani/datasets/refs/heads/main/fuel.csv", col_types = cols()) # reading the data
```

## Question 1: What is the combined miles per gallon by model year?

```{r}
year_summary <- fuel %>%
  group_by(year) %>% # group the dataset by year
  summarise(
    avg_mpg = mean(combined_mpg_ft1, na.rm = TRUE) # calculate the average of the combined miles per gallon that reflects a weighted average of both city and highway fuel economy ratings as avg_mpg
  ) %>%
  mutate( # make some new columns to calculate
    prev_mpg = lag(avg_mpg), # a column that returns the previous year's avg_mpg
    mpg_change = avg_mpg - prev_mpg, # a column that has the difference of the current year and previous year avg_mpg
    pct_change = (avg_mpg / prev_mpg - 1) * 100 # the percent change of that difference
  )

year_summary
```
```{r}
year_summary %>%
  filter(year > 2005) #noticing that 2008 is the start of an explosive trend of larger percent increases
```
```{r}
mean_pct_change <- year_summary %>%
  filter(year > 2007) %>%
  summarise(mean_pct_change = mean(pct_change, na.rm = TRUE))

mean_pct_change
```
```{r}
sd_pct_change <- year_summary %>%
  filter(year > 2007) %>%
  summarise(std_pct_change = sd(pct_change, na.rm = TRUE))

sd_pct_change
```
```{r}
sd_pct_change/mean_pct_change # tells us that the standard deviation of percent change of mpg after 2007 is 83% of the mean suggesting unstable rate of growth, not exactly exponential because it was otherwise be closer to 0% 
```


```{r}
# Pull specific values for annotations
start_val <- year_summary %>% filter(year == min(year)) # the first point of the graph's data, the earliest year's avg_mpg
end_val   <- year_summary %>% filter(year == max(year)) # the second point of the graph's data, the latest year's avg_mpg
```


```{r}
start_val
```


```{r}
end_val
```



```{r}
library(ggplot2)
library(plotly)

p <- year_summary %>%
  ggplot(aes(x = year, y = avg_mpg)) + # year on x, average combined mpg on y
  geom_line() + # line graph in black
  geom_smooth(se = FALSE) + # smoothed trend line in blue and removing the confidence interval
  geom_vline(xintercept = 2008, linetype = "dashed", color = "red") + # a dashed, red, vertical line at the year 2008
  annotate("text", x = 2003.5, y = 20.8, label = "Efficiency upturn of 2008 →",
           hjust = -0.05, size = 3, color = "gray30") +
  annotate("text", x = start_val$year, y = start_val$avg_mpg + 0.4,
           label = paste0(round(start_val$avg_mpg, 1), " MPG"), size = 3) +
  annotate("text", x = end_val$year + 1, y = end_val$avg_mpg - .25,
           label = paste0(round(end_val$avg_mpg, 1), " MPG"), size = 3) +
  theme_minimal() +
  labs(title = "Average combined MPG by model year",
       x = "Year",
       y = "Combined MPG")

ggplotly(p)
```

I chose to be more interested in the combined MPG metric because gas prices will fluctuate and inflate over time. I wanted to see the MPG increase over the years that cars were made, and you can see that barely right at 2008 of the both lines, where there is a turning point in the smoothed line like the right half of a "U" shape that you see in a graph of an exponential function, marking it as the point of the beginning of a positive trend of cars' gas efficiency. After 2007, there was an average yearly percentage increase of 2.65% with a standard deviation of 2.2%, indicating an unstable rate of positive growth, though we cannot call it exponential growth because the standard deviation of the rate is not close enough to zero.

## Question 2: How does Combined MPG perform for the different vehicle classes?

```{r}
library(plotly)
library(dplyr)

# Build summary stats per class for clean tooltips
class_stats <- fuel %>%
  filter(!is.na(class)) %>%
  group_by(class) %>%
  summarise(
    median_mpg = median(combined_mpg_ft1, na.rm = TRUE),
    q1 = quantile(combined_mpg_ft1, 0.25, na.rm = TRUE),
    q3 = quantile(combined_mpg_ft1, 0.75, na.rm = TRUE),
    min_mpg = min(combined_mpg_ft1, na.rm = TRUE),
    max_mpg = max(combined_mpg_ft1, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(median_mpg) # sort by median so order matches the plot


# Build one trace per class so tooltips are clean
plot_ly() %>%
  add_trace(
    data = fuel %>% filter(!is.na(class)),
    type = "box",
    x = ~combined_mpg_ft1,
    y = ~reorder(class, combined_mpg_ft1, median),
    orientation = "h",
    hovertemplate = paste(
      "<b>%{y}</b><br>",
      "Median: %{median}<br>",
      "Q1–Q3: %{q1}–%{q3}<br>",
      "Range: %{lowerfence}–%{upperfence}<extra></extra>"
    ),
    marker = list(color = "steelblue", opacity = 0.6),
    line = list(color = "steelblue")
  ) %>%

  layout(
    title = "Combined MPG by vehicle class",
    xaxis = list(title = "Combined MPG"),
    yaxis = list(title = ""),
    height = 1000,  # tall enough to give each class breathing room
    margin = list(l = 200),
    showlegend = TRUE,
    legend = list(orientation = "h", y = -0.08)
  )
```

